---
title: "MicroVAR Tutorial"
output: 
  learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(MicroVAR)
knitr::opts_chunk$set(echo = FALSE)
```

## 16s rRNA Analysis Preprocessing

### Data Preprocessing

Preprocessing of 16S rRNA (Amplicon) using QIIME2.\
Utilized the 2023.07 version.

```{bash, eval = FALSE, echo = TRUE}
# Perform denoising of paired-end sequences using DADA2
qiime dada2 denoise-paired \
--i-demultiplexed-seqs work/paired-end-demux.qza \
--p-trunc-len-f ${forward} \
--p-trunc-len-r ${reverse} \
--o-table work/table.qza \
--o-representative-sequences work/rep-seqs.qza \
--o-denoising-stats work/denoising-stats.qza

# Generate a phylogenetic tree using MAFFT and FastTree
qiime phylogeny align-to-tree-mafft-fasttree \
--i-sequences work/rep-seqs.qza \
--o-alignment work/aligned-rep-seqs.qza \
--o-masked-alignment work/masked-aligned-rep-seqs.qza \
--o-tree work/unrooted-tree.qza \
--o-rooted-tree work/rooted-tree.qza

# Create a directory for output files
mkdir output

# Export the OTU table to the output directory
qiime tools export \
--input-path work/table.qza \
--output-path output

# Convert the OTU table from biom format to tsv format
biom convert \
-i output/feature-table.biom \
-o output/otu_table.tsv \
--to-tsv

# Clean up the tsv file by removing the first line and modifying the header
cd output
sed -i '1d' otu_table.tsv
sed -i 's/#OTU ID//' otu_table.tsv
cd ../

# Export the representative sequences to the output directory
qiime tools export \
--input-path work/rep-seqs.qza \
--output-path output

# Export the unrooted tree file to the output directory
qiime tools export \
--input-path work/unrooted-tree.qza \
--output-path output
cd output
mv tree.nwk unrooted_tree.nwk
cd ../

# Export the rooted tree file to the output directory
qiime tools export \
--input-path work/rooted-tree.qza \
--output-path output
cd output
mv tree.nwk rooted_tree.nwk
cd ../
```

Preprocessing using QIIME2 is almost complete, except for the taxonomy classification process.\
Before proceeding with the final step, you need to set up the classifier. You can download and use classifiers provided at <https://resources.qiime2.org/>, or you can train one yourself using QIIME2's feature-classifier. Here, we used the reference data from Greengenes version 2022.10 and trained it using the Naive Bayes method. This classifier is also available at <https://resources.qiime2.org/>.

The following code classifies taxa using the classifier.

```{bash, eval = FALSE, echo = TRUE}
# Classify the sequences using a pre-trained classifier
qiime feature-classifier classify-sklearn \
  --i-reads work/rep-seqs.qza \
  --i-classifier ${classifier} \
  --o-classification work/taxonomy.qza

# Export the taxonomy classification results to the output directory
qiime tools export \
--input-path work/taxonomy.qza \
--output-path output
```

As an additional preprocessing step, functional prediction was performed using Picrust2

```{bash, eval = FALSE, echo = TRUE}
# Convert the OTU table from TSV format to BIOM format
biom convert -i output/otu_table.tsv -o work/asv_picrust.biom --table-type="OTU table" --to-json

# Run the PICRUSt2 pipeline for functional prediction
picrust2_pipeline.py -s output/dna-sequences.fasta -i work/asv_picrust.biom -o work/picrust_result/

# Copy the predicted metagenome results to the output directory
cp work/picrust_result/KO_metagenome_out/pred_metagenome_unstrat.tsv.gz output/picrust.tsv.gz
```

## Shotgun Metagenomics

### Data Preprocessing

We will perform sample preprocessing using FastQC, Trim Galore, Bowtie2, and Samtools, followed by reference-based taxonomy classification using Kraken2 and Bracken.

Sample preprocessing

```{bash, eval = FALSE, echo = TRUE}
samples_file="data/sample.txt"
samples=($(<"$samples_file"))

mkdir -p work/trim_seq

# Loop through each sample and run Trim Galore for paired-end trimming
for sample in "${samples[@]}"; do
    trim_galore --paired "data/${sample}_1.fastq" "data/${sample}_2.fastq" --output_dir work/trim_seq
done

cd work/trimseq

mkdir trimming_report 
mv *.txt trimming_report

# Script to remove host sequences using Bowtie2 and convert output to BAM format using Samtools
mkdir -p work/host_removed_seq

# Loop through each sample and run Bowtie2 for host sequence removal
for sample in "${samples[@]}"; do
    bowtie2 -p 15 -x DB/h19_index/hg19 \
    -1 "work/trim_seq/${sample}_1_val_1.fq" \
    -2 "work/trim_seq/${sample}_2_val_2.fq" \
    --very-sensitive-local \
    --un-conc-gz "work/host_removed_seq/${sample}_host_removed" \
    > "work/host_removed_seq/${sample}_mapped_and_unmapped.sam"

    # Rename the output files for host removed sequences
    mv "work/host_removed_seq/${sample}_host_removed.1" "work/host_removed_seq/${sample}_host_removed_1.fastq.gz"
    mv "work/host_removed_seq/${sample}_host_removed.2" "work/host_removed_seq/${sample}_host_removed_2.fastq.gz"
    
    # Convert SAM file to BAM format using Samtools and remove the original SAM file
    samtools view -bS work/host_removed_seq/${sample}_mapped_and_unmapped.sam > work/host_removed_seq/${sample}_mapped_and_unmapped.bam
    rm work/host_removed_seq/${sample}_mapped_and_unmapped.sam
done 
```

Taxonomy classification

```{bash, eval = FALSE, echo = TRUE}
(base) [kcy@snu command]$ cat 2_1_taxonomy_kraken.sh 
samples_file="data/sample.txt"
samples=($(<"$samples_file"))

# Set the database path for Kraken2
DB=DB/k2_standard_20231009

mkdir -p output/kraken

# Loop through each sample and run Kraken2 for taxonomy classification
for sample in "${samples[@]}"; do
    R1="work/host_removed_seq/${sample}_host_removed_1.fastq.gz"
    R2="work/host_removed_seq/${sample}_host_removed_2.fastq.gz"
    kraken2 \
        --db $DB \
        --threads 70 \
        --confidence 0.5 \
        --minimum-base-quality 22 \
        --report output/kraken/${sample}.report \
        --paired $R1 $R2 \
        > output/kraken/${sample}.txt
done

mkdir -p output/bracken

# Loop through each sample and run Bracken for abundance estimation
for sample in "${samples[@]}"; do
    K2_REPORT="output/kraken/${sample}.report"

    bracken \
        -d $DB \
        -i $K2_REPORT \
        -o output/bracken/${sample}.bracken \
        -w output/bracken/${sample}_bracken_species.report \
        -r 150 \
        -l S
done

# Convert Bracken reports to a BIOM file format
kraken-biom output/bracken/*.report --fmt json -o output/bracken/taxa.biom
```

## Data Analyses

With the preprocessed data, you can perform microbiome analysis using the MicroVAR package.\
- Data integration\
- Alpha diversity analysis\
- Beta diversity analysis\
- Taxonomy analysis\
- Differential abundance analysis

## Data integration

### Begin microbiome analysis using the MicroVAR package in R

Download and load the MicroVAR package

```{r, eval = FALSE, echo = TRUE}
# install.packages("devtools")
devtools::install_github("ERASMUSlab/MicroVAR")

library(MicroVAR)
```

### Creating Phyloseq Data

```{r, message=F, warning=F, error=T, echo = TRUE}
metadata_path <- system.file("extdata", "AmpliconMetadata.csv", package = "MicroVAR")
asv_path <- system.file("extdata", "AmpliconASV.tsv", package = "MicroVAR")
taxa_path <- system.file("extdata", "AmpliconTaxonomy.tsv", package = "MicroVAR")
tree_path <- system.file("extdata", "AmpliconRootedTree.nwk", package = "MicroVAR")

amplicon <<- createAmpPhylo(metadata_path = metadata_path,
                            asv_path = asv_path, 
                            taxa_path = taxa_path, 
                            tree_path = tree_path)
amplicon
```

```{r, message=F, warning=F, error=T, echo = TRUE}
metadata_path <- system.file("extdata", "ShotgunMetadata.csv", package = "MicroVAR")
biom_path <- system.file("extdata", "ShotgunBiomData.biom", package = "MicroVAR")

shotgun <<- createShotPhylo(biom_path = biom_path, 
                            metadata_path = metadata_path)
shotgun

```

## Alpha diversity analysis

### Amplicon data

```{r, message=F, warning=F, error=T, echo = TRUE, fig.width=5, fig.height=5}
shannon <- plotAlphaDiversity(phyloseq = amplicon, 
                              condition_col = "condition", 
                              condition_label = c("Fl","Fl_Arg"), 
                              colors = c('#FFA726','#26A69A'),
                              measure = "Shannon",
                              text_size = 15)
shannon

observed <- plotAlphaDiversity(phyloseq = amplicon,
                               condition_col = "condition", 
                               condition_label = c("Fl", "Fl_Arg"), 
                               colors = c('#FFA726','#26A69A'), 
                               measure ="Observed",
                               text_size = 15)
observed

simpson <- plotAlphaDiversity(phyloseq = amplicon,
                              condition_col = "condition",
                              condition_label = c("Fl", "Fl_Arg"),
                              colors = c('#FFA726','#26A69A'), 
                              measure ="Simpson",
                              text_size = 15)
simpson

chao1 <- plotAlphaDiversity(phyloseq = amplicon,  
                            condition_col = "condition", 
                            condition_label = c("Fl", "Fl_Arg"), 
                            colors = c('#FFA726','#26A69A'), 
                            measure ="Chao1",text_size = 15)

chao1 
```

### Shotgun data

```{r, message=F, warning=F, error=T, echo = TRUE, fig.width=5, fig.height=5}
shannon <- plotAlphaDiversity(phyloseq = shotgun, 
                              condition_col = "condition", 
                              condition_label = c("Fl","Fl_Arg"), 
                              colors = c('#FFA726','#26A69A'),
                              measure = "Shannon",
                              text_size = 15)
shannon

observed <- plotAlphaDiversity(phyloseq = shotgun,
                               condition_col = "condition", 
                               condition_label = c("Fl", "Fl_Arg"), 
                               colors = c('#FFA726','#26A69A'), 
                               measure ="Observed",
                               text_size = 15)
observed

simpson <- plotAlphaDiversity(phyloseq = shotgun,
                              condition_col = "condition",
                              condition_label = c("Fl", "Fl_Arg"),
                              colors = c('#FFA726','#26A69A'), 
                              measure ="Simpson",
                              text_size = 15)
simpson

chao1 <- plotAlphaDiversity(phyloseq = shotgun,  
                            condition_col = "condition", 
                            condition_label = c("Fl", "Fl_Arg"), 
                            colors = c('#FFA726','#26A69A'), 
                            measure ="Chao1",text_size = 15)

chao1 
```

## Beta diversity analysis

### Amplicon data

```{r, message=F, warning=F, error=T, echo = TRUE, fig.width=5, fig.height=5}
jaccard = plotBetaDiversity(phyloseq = amplicon,
                            condition_col = "condition", 
                            condition_label = c("Fl", "Fl_Arg"), 
                            colors = c('#FFA726','#26A69A'), 
                            measure ="jaccard",
                            text_size = 15)
jaccard

bray = plotBetaDiversity(phyloseq = amplicon, 
                         condition_col = "condition", 
                         condition_label = c("Fl", "Fl_Arg"), 
                         colors = c('#FFA726','#26A69A'), 
                         measure ="bray",
                         text_size = 15)
bray

wunifrac = plotBetaDiversity(phyloseq = amplicon,
                             condition_col = "condition", 
                             condition_label = c("Fl", "Fl_Arg"), 
                             colors = c('#FFA726','#26A69A'), 
                             measure ="wunifrac",
                             text_size = 15)
wunifrac

unifrac = plotBetaDiversity(phyloseq = amplicon, 
                            condition_col = "condition", 
                            condition_label = c("Fl", "Fl_Arg"), 
                            colors = c('#FFA726','#26A69A'), 
                            measure ="unifrac",
                            text_size = 15)
unifrac
```

### Shotgun data

```{r, message=F, warning=F, error=T, echo = TRUE, fig.width=5, fig.height=5}
jaccard = plotBetaDiversity(phyloseq = shotgun,
                            condition_col = "condition", 
                            condition_label = c("Fl", "Fl_Arg"), 
                            colors = c('#FFA726','#26A69A'), 
                            measure ="jaccard",
                            text_size = 15)
jaccard

bray = plotBetaDiversity(phyloseq = shotgun, 
                         condition_col = "condition", 
                         condition_label = c("Fl", "Fl_Arg"), 
                         colors = c('#FFA726','#26A69A'), 
                         measure ="bray",
                         text_size = 15)
bray

unifrac = plotBetaDiversity(phyloseq = shotgun, 
                            condition_col = "condition", 
                            condition_label = c("Fl", "Fl_Arg"), 
                            colors = c('#FFA726','#26A69A'), 
                            measure ="unifrac",
                            text_size = 15)
unifrac
```

## Taxonomy analysis

### Amplicon data

```{r, message=F, warning=F, error=T, echo = TRUE, fig.width=5, fig.height=5}

phylum_taxa <- plotTaxonomyBars(phyloseq = amplicon,
                                condition_label = c("Fl","Fl_Arg"),
                                taxa_level = "Phylum",
                                keep_percent = 1,
                                text_size = 10)
phylum_taxa

class_taxa <- plotTaxonomyBars(phyloseq = amplicon,
                                condition_label = c("Fl","Fl_Arg"),
                                taxa_level = "Class",
                                keep_percent = 1,
                                text_size = 10)
class_taxa

order_taxa <- plotTaxonomyBars(phyloseq = amplicon,
                                condition_label = c("Fl","Fl_Arg"),
                                taxa_level = "Order",
                                keep_percent = 1,
                                text_size = 10)
order_taxa

family_taxa <- plotTaxonomyBars(phyloseq = amplicon,
                                condition_label = c("Fl","Fl_Arg"),
                                taxa_level = "Family",
                                keep_percent = 1,
                                text_size = 7)
family_taxa

genus_taxa <- plotTaxonomyBars(phyloseq = amplicon,
                               condition_label = c("Fl","Fl_Arg"),
                               taxa_level = "Genus",
                               keep_percent = 1,
                               text_size = 7)
genus_taxa

species_taxa <- plotTaxonomyBars(phyloseq = amplicon,
                                 condition_label = c("Fl","Fl_Arg"),
                                 taxa_level = "Species",
                                 keep_percent = 1,
                                 text_size = 10)
species_taxa
```

### Shotgun data

```{r, message=F, warning=F, error=T, echo = TRUE, fig.width=5, fig.height=5}

phylum_taxa <- plotTaxonomyBars(phyloseq = shotgun,
                                condition_label = c("Fl","Fl_Arg"),
                                taxa_level = "Phylum",
                                keep_percent = 1,
                                text_size = 10)
phylum_taxa

class_taxa <- plotTaxonomyBars(phyloseq = shotgun,
                                condition_label = c("Fl","Fl_Arg"),
                                taxa_level = "Class",
                                keep_percent = 1,
                                text_size = 10)
class_taxa

order_taxa <- plotTaxonomyBars(phyloseq = shotgun,
                                condition_label = c("Fl","Fl_Arg"),
                                taxa_level = "Order",
                                keep_percent = 1,
                                text_size = 10)
order_taxa

family_taxa <- plotTaxonomyBars(phyloseq = shotgun,
                                condition_label = c("Fl","Fl_Arg"),
                                taxa_level = "Family",
                                keep_percent = 1,
                                text_size = 10)
family_taxa

genus_taxa <- plotTaxonomyBars(phyloseq = shotgun,
                               condition_label = c("Fl","Fl_Arg"),
                               taxa_level = "Genus",
                               keep_percent = 1,
                               text_size = 7)
genus_taxa

species_taxa <- plotTaxonomyBars(phyloseq = amplicon,
                                 condition_label = c("Fl","Fl_Arg"),
                                 taxa_level = "Species",
                                 keep_percent = 1,
                                 text_size = 6)
species_taxa
```

## Differential abundance analysis

### Amplicon data

```{r, message=F, warning=F, error=T, echo = TRUE, fig.width=20, fig.height=10, out.width='100%', out.height='100%'}
plotHeatTree(phyloseq = amplicon,
             taxa_p = 0.1,
             seed = 0,
             p_adjust = "BH",
             group ="condition",
             condition_label = c("Fl", "Fl_Arg"),
             pvalue_cutoff = 0.05, 
             norm = "TSS",
             clade_label_level = 7,
             lade_label_font_size = 10,
             colors = c('#FFA726','#26A69A'),
             text_size = 10)
```

### Shotgun data

```{r, message=F, warning=F, error=T, echo = TRUE, fig.width=20, fig.height=10, out.width='100%', out.height='100%'}
plotHeatTree(phyloseq = shotgun,
             taxa_p = 0.1,
             seed = 0,
             p_adjust = "BH",
             group ="condition",
             condition_label = c("Fl", "Fl_Arg"),
             pvalue_cutoff = 0.05, 
             norm = "TSS",
             clade_label_level = 7,
             lade_label_font_size = 10,
             colors = c('#FFA726','#26A69A'),
             text_size = 10)
```
